### 2025-05-08 chapters/arima2/child.tex
Added draft for ARMA chapter

<<armaSetup, include=FALSE,echo=FALSE,results='hide'>>=

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  dev = 'cairo_pdf'
)

library(arima2)      # Package described in this article
library(tidyverse)   # Data wrangling and visualizations
library(foreach)     # Parallel Computing
library(doParallel)  # Parallel Computing
library(doRNG)       # To get reproducible seed for parallel computing
library(pomp)        # Used for the bake function
library(knitr)       # Necessary for Rnw file
library(latex2exp)

# Setting black and white ggplot2 theme for entire document.
theme_set(
  theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5)
    )
)

AR_labs <- function(num) {
  paste0("AR = ", num)
}

MA_labs <- function(num) {
  paste0("MA = ", num)
}
@

@@

<<setupMulti>>=
sample_data <- function(data, job, n, P, Q, ...) {

  res <- list()

  coefs <- sample_ARMA_coef(
    order = c(P, Q), Mod_bounds = c(0.1, 0.9), min_inv_root_dist = 0.1
  ) |> round(digits = 2)

  res$ars <- coefs[grepl("^ar[[:digit:]]+", names(coefs))]
  res$mas <- coefs[grepl("^ma[[:digit:]]+", names(coefs))]

  res$x <- arima.sim(
    n = n,
    model = list(ma = res$mas)
  )

  res
}

checkImprove <- function(data, job, instance, ...) {

  arma2_fit <- tryCatch(
    arima2::arima(
      instance$x, order = c(0, 0, 1), SSinit = 'Rossignol2011', max_inv_root = 0.99,
      max_repeats = 10, include.mean = FALSE
    ),
    error = function(e) list(num_starts = 0, coef = c('ma1' = 0))
  )

  c('num_starts' = arma2_fit$num_starts, 'll_improve' = ifelse(arma2_fit$num_starts > 10, arma2_fit$all_values[arma2_fit$num_starts] - arma2_fit$all_values[1], NA_real_), "truth" = instance$mas, "est" = arma2_fit$coef)
}

get_profile <- function(model, npts) {
  results <- matrix(ncol = 2, nrow = npts)

  fit_arma <- function(fixed) {

    P <- model$arma[1]
    Q <- model$arma[2]

    arima2::arima(
      model$x, order = c(0, 0, 1), fixed = c(fixed), SSinit = 'Rossignol2011',
      max_iters = 1, include.mean = FALSE
    )$loglik
  }

  fixed_vals <- seq(-1+1e-8, 1-1e-8, length.out = npts)
  lls <- purrr::map_dbl(fixed_vals, fit_arma)
  results[, 1] <- fixed_vals
  # results[, 2] <- model$coef[2]
  results[, 2] <- lls
  colnames(results) <- c(names(model$coef), 'loglik')
  as.data.frame(results)
}

ma1_res <- readRDS("data/sim_MA1_results.rds")

ma1_improve <- ma1_res %>%
  filter(num_starts > 10) %>%
  filter(abs(est.ma1) < 0.95)

exp_seed <- 100000
prob_seed <- 50000

# Loop --------------------------------------------------------------------

job_ids <- c(10463, 16385, 22383, 41307)
ggs <- list()

for (j in 1:length(job_ids)) {
  i <- job_ids[j]
  # i <- test_mods[j]
  set.seed(prob_seed + (i %% 25000) - 1)

  instance <- sample_data(
    NULL,
    i,
    pull(ma1_res, n)[i],
    P = 0, Q = 1
  )

  set.seed(exp_seed + i)

  arma2_fit <- tryCatch(
    arima2::arima(
      instance$x, order = c(0, 0, 1), SSinit = 'Rossignol2011', max_inv_root = 0.99,
      max_repeats = 10, include.mean = FALSE
    ),
    error = function(e) list(num_starts = 0)
  )

  stats1 <- stats::arima(
    instance$x,
    order = c(0, 0, 1),
    SSinit = 'Rossignol2011',
    include.mean = FALSE
  )

  stats2 <- stats::arima(
    instance$x,
    order = c(0, 0, 1),
    SSinit = 'Rossignol2011',
    include.mean = FALSE,
    method = 'CSS'
  )

  mod_prof <- get_profile(arma2_fit, 1000)
  # mod_prof <- profile(arma2_fit, which = 1, lower = -1 + 1e-8, upper = 1 - 1e-8, npts = 1000)

  diffs <- case_when(
    j == 1 ~ 0.9,
    j == 2 ~ 3,
    j == 3 ~ 0.5,
    j == 4 ~ 0.5,
    TRUE ~ 2
  )

  df_lines <- data.frame(
    'value' = c(
      instance$mas,
      arma2_fit$coef[1],
      stats1$coef[1],
      stats2$coef[1]
    ),
    'type1' = c('truth', 'est', 'est', 'init'),
    'type2' = c('truth', 'est1', 'est2', 'truth')
  )

  gg_tmp <- mod_prof %>%
    filter(loglik > max(loglik) - diffs) %>%
    ggplot(aes(x = ma1, y = loglik)) +
    geom_line() +
    geom_vline(data = df_lines, aes(xintercept = value, col = type2, linetype = type1)) +
    geom_vline(xintercept = instance$mas, col = 'black') +
    geom_vline(xintercept = stats2$coef, col = 'black', linetype = 'dotted') +
    theme_bw() +
    labs(
      x = TeX('$\\theta_1$'),
      y = "Loglikelihood"
    ) +
    theme(plot.title = element_text(hjust = 0.5)) +
    # Add manual legend using dummy data
    scale_linetype_manual(
      name = NULL,
      values = c(
        "truth" = "solid",
        "est" = "dashed",
        "init" = "dotted"
      ), labels = c("truth" = 'Truth', "est" = "Estimate", "init" = 'CSS Initialization')
    ) +
    scale_color_manual(
      name = NULL,
      values = setNames(c("blue", "red", "black"), c("est1", "est2")),
      labels = setNames(c("arima2::arima", "stats::arima"), c("est1", "est2"))
    ) +
    guides(
      linetype = guide_legend(order = 1),
      color = guide_legend(order = 2)
    ) +
    theme(legend.title = element_blank(), legend.margin = margin(c(-2,0,0,0)),
          legend.key.spacing.y = unit(10, "pt"), legend.spacing.y = unit(35, 'pt'))

  if (j == 2 | j == 4) {
    gg_tmp <- gg_tmp + theme(axis.title.y = element_blank())
  }

  if (j == 1 | j == 2) {
    gg_tmp <- gg_tmp + theme(axis.title.x = element_blank())
  }

  ggs[[j]] <- gg_tmp
}
@

\begin{figure}[ht]
<<mutliMode-fig, fig.height=3.5, fig.width=7>>=
ggpubr::ggarrange(
  plotlist = ggs,
  legend = 'right',
  common.legend = TRUE
)
@
\caption{\label{fig:multiMode}The profile log-likelihood of data simulated from four distinct MA(1) models, demonstrating a few examples of multimodal likelihood surfaces. The solid, black line indicates the true value of $\theta_1$; the dotted line is the CSS-initialization. The dashed lines correspond to the estimate $\hat{\theta}_1$ using \code{stats:arima} (red) and our proposed algorithm (implemented in \code{arima2::arima}, blue).}
\end{figure}

@@

<<loadSimulation>>=
simulation1 <- readRDS("data/sim1_results.rds")

results_stats <- simulation1 %>%
  filter(algorithm == 'statsArima')

results_arima2 <- simulation1 %>%
  filter(algorithm == 'Arima2') %>%
  mutate(job.id = job.id - 36000)

wide_res <- dplyr::full_join(
  x = results_stats,
  y = results_arima2,
  by = c("job.id"),
  suffix = c("_arima", "_arima2")
) %>%
  select(
    job.id, P = P_arima, Q = Q_arima, n = n_arima, ll_arima,
    ll_arima2, err_arima, err_arima2, num_restarts_arima,
    num_restarts_arima2, n_covered_arima, n_covered_arima2,
    n_prof_covered_arima, n_prof_covered_arima2,
    n_mle_failures_arima, n_mle_failures_arima2, n_prof_failures_arima,
    n_prof_failures_arima2
  )

tot_pct <- wide_res %>%
  filter(!is.na(ll_arima2) & !is.na(ll_arima)) %>%
  summarize(pct = sum(ll_arima2 > ll_arima) / n()) %>%
  as.numeric()

arma33_n50_pct <- wide_res %>%
  filter(n == 50, P == 3, Q == 3) %>%
  mutate(ll_diff = ll_arima2 - ll_arima) %>%
  summarize(pct = sum(ll_diff > 0, na.rm = TRUE) / sum(!is.na(ll_arima2) & !is.na(ll_arima))) %>%
  as.numeric()

ARMA_pct <- wide_res %>%
  filter(!is.na(ll_arima2) & !is.na(ll_arima)) %>%
  filter(P != 0 & Q != 0) %>%
  summarize(pct = sum(ll_arima2 > ll_arima) / n()) %>%
  as.numeric()

diff_IQR <- wide_res %>%
  filter(ll_arima2 > ll_arima) %>%
  mutate(ll_diff = ll_arima2 - ll_arima) %>% pull(ll_diff) %>% quantile(probs = c(0.25, 0.75))

median_diff <- wide_res %>%
  filter(ll_arima2 > ll_arima) %>%
  mutate(ll_diff = ll_arima2 - ll_arima) %>% pull(ll_diff) %>%
  median()

mean_restarts <- wide_res %>%
  filter(ll_arima2 > ll_arima) %>%
  pull(num_restarts_arima2) %>%
  mean()
@

@@

Even in this extremely simplified scenario, existing software failed to properly maximize model likelihoods in at least $\Sexpr{round(100 * tot_pct, 1)}\%$ of the simulated datasets---evidenced by an improvement obtained using Algorithm~1.
Though this improvement may appear modest, an improvement in $\Sexpr{round(100 * tot_pct, 1)}\%$ of the large number of published ARMA models would affect many papers---a number measured in thousands of papers since 2024 following our estimate in the introduction.
Furthermore, time series analysis courses and textbooks often recommend fitting multiple ARMA models to a dataset, and here we only fit one for each algorithm. 
Consequently, the probability that at least one candidate model is not properly optimized increases significantly in practice.
The rate of improvement obtained using our algorithm increases with model complexity and decreases with more observations (Fig.~\ref{fig:simProps}).
For example, likelihoods improved in $\Sexpr{round(100 * arma33_n50_pct, 1)}\%$ of the simulations when $p = q = 3$ and $n = 50$; models of this size and number of observations are not uncommon in published research studies.

<<getTiming>>=
sim_times <- readRDS("data/sim1_timing.rds")
mean_time <- sim_times$time.running %>% as.numeric() %>% mean()
@

\begin{figure}[!ht]
<<simProps, fig.height=3, fig.width=4.5>>=
wide_res %>%
  filter(P != 0 & Q !=0) %>%
  group_by(P, Q, n) %>%
  summarize(pct_improve = sum(ll_arima2 > ll_arima, na.rm = TRUE) / n()) %>%
  ggplot(aes(x = n, y = pct_improve)) +
  facet_grid(Q~P, labeller = labeller(P = AR_labs, Q = MA_labs)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(limits = c(0, 0.6)) +
  ylab("Proportion of Improved Fits") +
  xlab("Number of observations (n)")
@
\caption{\label{fig:simProps}Proportion of simulated data with improved likelihood from using multiple restarts (Algorithm~\ref{alg:mle}).}
\end{figure}

Importantly, we do not claim to improve likelihood for all ARMA models and datasets.
In this simulation study, our likelihood maximum routine did not improve likelihoods for many of the simulated data.
However, these results demonstrate that there is a very real potential for obtaining sub-optimal parameter estimates when using only a single parameter initialization, even in the most idealistic scenarios.
Rather than having to worry if a single initialization is sufficient to fit a given model, it is preferable to adopt methods that make such situations rare.
The primary limitation of our algorithm is that the potential for improved fits comes at the cost of increased computation times.
In our simulation study, however, the average time to estimate parameters using our approach was $\Sexpr{round(mean_time, 1)}$ seconds, a computational expense that is worth the effort in many situations.

The median log-likelihood improvement in this simulation study was $\Sexpr{round(median_diff, 2)}$, with an interquartile range of $\Sexpr{paste0("(", paste0(round(diff_IQR, 2), collapse = ', '), ")")}$.
Among the most common motivations for fitting ARMA models is to model serial correlations in a regression model; in this setting, the discovered shortcomings in log-likelihood are often enough to change the outcome of the analysis.
For instance, consider modeling $y_i = \beta x_i + \epsilon_i$, where $\beta \in \mathbb{R}$, and we model the error terms $\epsilon_i \sim \text{ARMA}(p, q)$.
For now, we will assume the order $(p, q)$ is fixed.
We may wish to test the hypothesis $H_0: \beta = 0$ vs $H_1: \beta \neq 0$.
A standard approach to doing this is a likelihood ratio test, and using Wilks' theorem to get an approximate test.
We denote $ll_0$ and $ll_1$ as the maximum log-likelihood of the model under $H_0$ and $H_1$, respectively.
The standard approximation is to assume $2\Delta = 2(ll_1 - ll_0) \sim \chi^2_1$.
Using a significance level of $\alpha = 0.05$, we would reject $H_0$ if $\Delta \geq 1.92$.
Given that $E_{H_0}[\Delta] = 0.5$, subtracting the reported log-likelihood deficiencies (which has a median value of $\Sexpr{round(median_diff, 2)}$) of existing software to either or both $ll_0, ll_1$ could change the outcome of this test.

@@

\begin{figure}[!ht]
<<pctCovered, fig.height=4.2, fig.width=5.5>>=
gg_fisher <- simulation1 %>%
  filter(algorithm == "Arima2") %>%
  filter(P > 0 & Q > 0) %>%
  filter(n_mle_failures < 5, n_prof_failures < 5) %>%
  mutate(pct = n_covered / 100) %>%
  ggplot(aes(x = as.factor(n), y = pct, group = as.factor(n))) +
  geom_boxplot(outlier.stroke = 0.05) +
  facet_grid(P~Q, labeller = labeller(P = AR_labs, Q = MA_labs)) +
  geom_hline(yintercept = 0.95, linetype = 'dashed') +
  ylab("Coverage Proportion") +
  xlab("Number of Observations (n)") +
  theme(
    legend.title = element_blank(), legend.position = 'bottom',
    strip.background.y = element_blank(),
    strip.text.y = element_blank(),
    plot.margin = margin(20, 2, 5.5, 5.5),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 7)
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

gg_profile <- simulation1 %>%
  filter(algorithm == "Arima2") %>%
  filter(P > 0 & Q > 0) %>%
  filter(n_mle_failures < 5, n_prof_failures < 5) %>%
  mutate(pct = n_prof_covered / 100) %>%
  ggplot(aes(x = as.factor(n), y = pct, group = as.factor(n))) +
  geom_boxplot(outlier.stroke = 0.05) +
  facet_grid(P~Q, labeller = labeller(P = AR_labs, Q = MA_labs)) +
  geom_hline(yintercept = 0.95, linetype = 'dashed') +
  ylab("Coverage Proportion") +
  xlab("Number of Observations (n)") +
  theme(
    legend.position = 'none',
    axis.text.y = element_blank(), axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(20, 5.5, 5.5, 2),
    axis.text.x = element_text(size = 7)
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

ggpubr::ggarrange(
  gg_fisher, gg_profile, align = 'h', nrow = 1,
  labels = 'AUTO',
  label.x = c(0.16, 0)
)
@
\caption{\label{fig:pctCovered}Proportion of models that achieved nominal coverage of Bonferroni adjusted $95\%$ confidence intervals. The dashed line denotes the target coverage level. (A) Confidence intervals created using Fisher's information matrix. (B) Confidence intervals created using profile likelihoods.}
\end{figure}

@@

<<simTables>>=
simTables <- readRDS("data/sim2_AICtable_consistency.rds")
arma_inc <- simTables %>%
  filter(max_repeat == 1) %>%
  summarize(1 - sum(consistent, na.rm = TRUE) / n()) %>%
  unlist()
@

@@

The single parameter initialization approach resulted in AIC table inconsistencies in $\Sexpr{round(100*arma_inc, 1)}\%$ of the simulated datasets.
Although our proposed algorithm significantly mitigates this issue, it does not guarantee that all model likelihoods are fully maximized.
This is illustrated in Fig~\ref{fig:AICtabRes}, where a non-zero percentage of AIC tables remain inconsistent, even as the algorithm's stopping criterion grows.
The ARMA$(1, 1)$ panel in Fig~\ref{fig:AICtabRes} illustrates the increasing difficulty of parameter estimation when dealing with parameter redundancies.
In such cases, it is often necessary to adjust additional parameters in the numeric optimization routine.
For example, our R implementation of the algorithm relies on the generic BFGS optimizer in the \code{stats::optim} function.
Modifying the optimization method or the default hyperparameters can lead to improved fits or faster convergence rates.
While the default parameters of the numeric optimizer are generally adequate, increasing the maximum number of algorithmic iterations can be beneficial for fully maximizing the likelihood for challenging models and data.

\begin{figure}[!ht]
<<AICtabRes, fig.height=4.25, fig.width=6>>=
restart_labels <- c(
 '1'  = "M = 1",
 '10' = "M = 10",
 '30' = "M = 30",
 '50' = "M = 50"
)

simTables %>%
  group_by(n, P, Q, max_repeat) %>%
  summarize(
    pct_consistent = sum(consistent, na.rm = TRUE) / sum(!is.na(consistent))
  ) %>%
  ggplot(
    aes(
      x = n, y = pct_consistent,
      group = as.factor(max_repeat),
      color = as.factor(max_repeat),
      linetype = as.factor(max_repeat)
    )
  ) +
  geom_line() +
  geom_point() +
  facet_grid(P~Q, labeller = labeller(P = AR_labs, Q = MA_labs)) +
  ylab("Proportion of Tables\n that are Consistent") +
  xlab("Number of observations") +
  theme(legend.title = element_blank(), legend.position = 'bottom') +
  scale_color_discrete(labels = restart_labels) +
  scale_linetype_discrete(labels = restart_labels) +
  scale_y_continuous(limits = c(0.35, 1))
@
\caption{\label{fig:AICtabRes}Data is generated from ARMA$(p, q)$ models with $(p, q) \in \{1, 2, 3\}^2$, and the corresponding AIC table is created. The Y-axis shows the percentage of tables that were consistent. M is the number of times a maxima is observed before the algorithm terminates, so $M=1$ corresponds to the standard maximization procedure.}
\end{figure}

@@

\begin{figure}[!ht]
<<huron, fig.height=2.5,fig.width=4.5>>=
data("miHuron_level")
ggplot(miHuron_level, aes(x = Date, y = Average)) +
  geom_line() +
  ylab("Average Lake Depth (ft)") +
  theme(axis.title.x = element_blank())
@
\caption{\label{fig:huron}
Average depth of Lake Michigan-Huron from 1860-2014.
}
\end{figure}

@@

<<getTableValues, include=FALSE, warning=FALSE, cache=TRUE>>=
set.seed(12345)
P <- 3
Q <- 3
tab_tol <- 1e-2
huron_tab_vals <- aicTable(
  miHuron_level$Average, P = P, Q = Q, max_iters = 1,
  optim.control = list(maxit = 1000)
)
huron_tab_vals2 <- aicTable(
  miHuron_level$Average, P = P, Q = Q,
  optim.control = list(maxit = 1000)
)

is_improved <- huron_tab_vals2 + tab_tol < huron_tab_vals
@

\begin{table}[h]
    \centering
    % Start the first minipage for Table 1
    \begin{subtable}{0.4\textwidth}
        \centering
        \caption{\label{tab:AIC1}Single parameter initialization.}
        \begin{tabular}{c|c|c|c|c}\hline
            & MA0 & MA1 & MA2 & MA3 \\
            \hline
            AR0 & \Sexpr{ifelse(is_improved[1, 1], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[1, 1], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[1, 2], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[1, 2], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[1, 3], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[1, 3], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[1, 4], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[1, 4], 1), nsmall = 1)} \\
            \hline
            AR1 & \Sexpr{ifelse(is_improved[2, 1], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[2, 1], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[2, 2], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[2, 2], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[2, 3], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[2, 3], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[2, 4], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[2, 4], 1), nsmall = 1)} \\
            \hline
            AR2 & \Sexpr{ifelse(is_improved[3, 1], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[3, 1], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[3, 2], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[3, 2], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[3, 3], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[3, 3], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[3, 4], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[3, 4], 1), nsmall = 1)} \\
            \hline
            AR3 & \Sexpr{ifelse(is_improved[4, 1], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[4, 1], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[4, 2], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[4, 2], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[4, 3], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[4, 3], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[4, 4], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals[4, 4], 1), nsmall = 1)} \\
            \hline
        \end{tabular}
    \end{subtable}
    \hspace{15mm} % this fills the space between minipages if needed
    % Start the second minipage for Table 2
        \begin{subtable}{0.4\textwidth}
        \centering
        \caption{\label{tab:AIC2}Multiple parameter initializations.}
        \begin{tabular}{c|c|c|c|c}\hline
            & MA0 & MA1 & MA2 & MA3 \\
            \hline
            AR0 & \Sexpr{ifelse(is_improved[1, 1], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[1, 1], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[1, 2], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[1, 2], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[1, 3], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[1, 3], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[1, 4], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[1, 4], 1), nsmall = 1)} \\
            \hline
            AR1 & \Sexpr{ifelse(is_improved[2, 1], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[2, 1], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[2, 2], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[2, 2], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[2, 3], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[2, 3], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[2, 4], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[2, 4], 1), nsmall = 1)} \\
            \hline
            AR2 & \Sexpr{ifelse(is_improved[3, 1], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[3, 1], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[3, 2], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[3, 2], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[3, 3], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[3, 3], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[3, 4], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[3, 4], 1), nsmall = 1)} \\
            \hline
            AR3 & \Sexpr{ifelse(is_improved[4, 1], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[4, 1], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[4, 2], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[4, 2], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[4, 3], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[4, 3], 1), nsmall = 1)} & \Sexpr{ifelse(is_improved[4, 4], "\\improvedCell", "")} \Sexpr{format(round(huron_tab_vals2[4, 4], 1), nsmall = 1)} \\
            \hline
        \end{tabular}
    \end{subtable}

>>>>

@@

\begin{figure}[ht]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

{\centering \includegraphics[width=\maxwidth]{figure/mutliMode-fig-1}

}

\end{knitrout}
\caption{\label{fig:multiMode}The profile log-likelihood of data simulated from four distinct MA(1) models, demonstrating a few examples of multimodal likelihood surfaces. The solid, black line indicates the true value of $\theta_1$; the dotted line is the CSS-initialization. The dashed lines correspond to the estimate $\hat{\theta}_1$ using \code{stats:arima} (red) and our proposed algorithm (implemented in \code{arima2::arima}, blue).}
\end{figure}

@@

@@

Even in this extremely simplified scenario, existing software failed to properly maximize model likelihoods in at least $20.8\%$ of the simulated datasets---evidenced by an improvement obtained using Algorithm~1.
Though this improvement may appear modest, an improvement in $20.8\%$ of the large number of published ARMA models would affect many papers---a number measured in thousands of papers since 2024 following our estimate in the introduction.
Furthermore, time series analysis courses and textbooks often recommend fitting multiple ARMA models to a dataset, and here we only fit one for each algorithm. 
Consequently, the probability that at least one candidate model is not properly optimized increases significantly in practice.
The rate of improvement obtained using our algorithm increases with model complexity and decreases with more observations (Fig.~\ref{fig:simProps}).
For example, likelihoods improved in $55.1\%$ of the simulations when $p = q = 3$ and $n = 50$; models of this size and number of observations are not uncommon in published research studies.

\begin{figure}[!ht]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

{\centering \includegraphics[width=\maxwidth]{figure/simProps-1}

}

\end{knitrout}
\caption{\label{fig:simProps}Proportion of simulated data with improved likelihood from using multiple restarts (Algorithm~\ref{alg:mle}).}
\end{figure}

Importantly, we do not claim to improve likelihood for all ARMA models and datasets.
In this simulation study, our likelihood maximum routine did not improve likelihoods for many of the simulated data.
However, these results demonstrate that there is a very real potential for obtaining sub-optimal parameter estimates when using only a single parameter initialization, even in the most idealistic scenarios.
Rather than having to worry if a single initialization is sufficient to fit a given model, it is preferable to adopt methods that make such situations rare.
The primary limitation of our algorithm is that the potential for improved fits comes at the cost of increased computation times.
In our simulation study, however, the average time to estimate parameters using our approach was $0.6$ seconds, a computational expense that is worth the effort in many situations.

The median log-likelihood improvement in this simulation study was $0.66$, with an interquartile range of $(0.22, 1.46)$.
Among the most common motivations for fitting ARMA models is to model serial correlations in a regression model; in this setting, the discovered shortcomings in log-likelihood are often enough to change the outcome of the analysis.
For instance, consider modeling $y_i = \beta x_i + \epsilon_i$, where $\beta \in \mathbb{R}$, and we model the error terms $\epsilon_i \sim \text{ARMA}(p, q)$.
For now, we will assume the order $(p, q)$ is fixed.
We may wish to test the hypothesis $H_0: \beta = 0$ vs $H_1: \beta \neq 0$.
A standard approach to doing this is a likelihood ratio test, and using Wilks' theorem to get an approximate test.
We denote $ll_0$ and $ll_1$ as the maximum log-likelihood of the model under $H_0$ and $H_1$, respectively.
The standard approximation is to assume $2\Delta = 2(ll_1 - ll_0) \sim \chi^2_1$.
Using a significance level of $\alpha = 0.05$, we would reject $H_0$ if $\Delta \geq 1.92$.
Given that $E_{H_0}[\Delta] = 0.5$, subtracting the reported log-likelihood deficiencies (which has a median value of $0.66$) of existing software to either or both $ll_0, ll_1$ could change the outcome of this test.

@@

\begin{figure}[!ht]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

{\centering \includegraphics[width=\maxwidth]{figure/pctCovered-1}

}

\end{knitrout}
\caption{\label{fig:pctCovered}Proportion of models that achieved nominal coverage of Bonferroni adjusted $95\%$ confidence intervals. The dashed line denotes the target coverage level. (A) Confidence intervals created using Fisher's information matrix. (B) Confidence intervals created using profile likelihoods.}
\end{figure}

@@

@@

The single parameter initialization approach resulted in AIC table inconsistencies in $45.6\%$ of the simulated datasets.
Although our proposed algorithm significantly mitigates this issue, it does not guarantee that all model likelihoods are fully maximized.
This is illustrated in Fig~\ref{fig:AICtabRes}, where a non-zero percentage of AIC tables remain inconsistent, even as the algorithm's stopping criterion grows.
The ARMA$(1, 1)$ panel in Fig~\ref{fig:AICtabRes} illustrates the increasing difficulty of parameter estimation when dealing with parameter redundancies.
In such cases, it is often necessary to adjust additional parameters in the numeric optimization routine.
For example, our R implementation of the algorithm relies on the generic BFGS optimizer in the \code{stats::optim} function.
Modifying the optimization method or the default hyperparameters can lead to improved fits or faster convergence rates.
While the default parameters of the numeric optimizer are generally adequate, increasing the maximum number of algorithmic iterations can be beneficial for fully maximizing the likelihood for challenging models and data.

\begin{figure}[!ht]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

{\centering \includegraphics[width=\maxwidth]{figure/AICtabRes-1}

}

\end{knitrout}
\caption{\label{fig:AICtabRes}Data is generated from ARMA$(p, q)$ models with $(p, q) \in \{1, 2, 3\}^2$, and the corresponding AIC table is created. The Y-axis shows the percentage of tables that were consistent. M is the number of times a maxima is observed before the algorithm terminates, so $M=1$ corresponds to the standard maximization procedure.}
\end{figure}

@@

\begin{figure}[!ht]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

{\centering \includegraphics[width=\maxwidth]{figure/huron-1}

}

\end{knitrout}
\caption{\label{fig:huron}
Average depth of Lake Michigan-Huron from 1860-2014.
}
\end{figure}

@@

\begin{table}[h]
    \centering
    % Start the first minipage for Table 1
    \begin{subtable}{0.4\textwidth}
        \centering
        \caption{\label{tab:AIC1}Single parameter initialization.}
        \begin{tabular}{c|c|c|c|c}\hline
            & MA0 & MA1 & MA2 & MA3 \\
            \hline
            AR0 &  166.8 &  46.6 &  7.3 &  -15.0 \\
            \hline
            AR1 &  -38.0 &  -37.4 &  -35.5 &  -33.8 \\
            \hline
            AR2 &  -37.3 &  -38.4 &  -36.9 &  -34.9 \\
            \hline
            AR3 &  -35.5 & \improvedCell -35.2 & \improvedCell -33.0 & \improvedCell -33.3 \\
            \hline
        \end{tabular}
    \end{subtable}
    \hspace{15mm} % this fills the space between minipages if needed
    % Start the second minipage for Table 2
        \begin{subtable}{0.4\textwidth}
        \centering
        \caption{\label{tab:AIC2}Multiple parameter initializations.}
        \begin{tabular}{c|c|c|c|c}\hline
            & MA0 & MA1 & MA2 & MA3 \\
            \hline
            AR0 &  166.8 &  46.6 &  7.3 &  -15.0 \\
            \hline
            AR1 &  -38.0 &  -37.4 &  -35.5 &  -33.8 \\
            \hline
            AR2 &  -37.3 &  -38.4 &  -36.9 &  -34.9 \\
            \hline
            AR3 &  -35.5 & \improvedCell -36.9 & \improvedCell -36.4 & \improvedCell -36.2 \\
            \hline
        \end{tabular}
    \end{subtable}
    \caption{\label{tab:huronTab}AIC values for an ARMA(p, q) model fit to Lake Michigan-Huron depths. Table~\ref{tab:AIC1} was computing using only a single parameter initialization. Table~\ref{tab:AIC2} was computed using Algorithm~\ref{alg:mle}. Highlighted cells show where the likelihood was improved (AIC reduced) using our algorithm.}
\end{table}

Recall that the AIC is defined as:
\begin{equation}
  \text{AIC} = -2 \max_{\allVar} \ell(\allVar; x^*) + 2d,\label{eq:AIC}
\end{equation}
where $\ell(\allVar; x^*)$ denotes the log-likelihood of a model indexed by parameter vector $\allVar \in \mathbb{R}^d$, $d \geq 1$, given the observed data $x^*$.
In the case of an ARMA model with an intercept, $d = p + q + 2$, where the additional parameter corresponds to a variance estimate.
If either $p$ or $q$ increases by one, then a corresponding increase in AIC values greater than two suggests that the \emph{inclusion} of an additional parameter resulted in a \emph{decrease} in the maximum of the log-likelihood, which is mathematically impossible under proper optimization.
Several such cases are present in Table~\ref{tab:AIC1}, for example increasing from an $\text{ARMA}(2, 2)$ model to a $\text{ARMA}(3, 2)$ model results in
a decrease of 1.0 log-likelihood units.
In this case, using our multiple restart algorithm eliminates all instances of mathematical inconsistencies (Table~\ref{tab:AIC2}).
We refer to tables that have log-likelihood values larger for any smaller nested model within the table as \emph{inconsistent}.
